---
title: "Introduction to R | SoSS PGR Workshop"
author: "Dr Pete McKenna"
date: "2024-10-02"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


# Rationale

- WEIRD datasets more commonly available and do not enhance R teaching equity
- Use of non-WERID datasets may serve to promote feelings of belonging and inclusivity in diverse student samples
- Students implicitly learn about the challenges of dataset retrieval and the importance of good metadata


# Data

- I have found data from 'Global Gender Gap Report 2024' for Sub-Saharan Africa
- The goal is to read and interpret this data in R
- The data contains a iso3 code vector, so I've also found a related dataset to provide country name data. 


# Load packages

These will need to be installed using the console if not already installed. 

```{r}

# Load the tidyverse
library(tidyverse)
library(psych)
library(skimr)

```


# Read in the data

```{r}

# Read in the GGGI data
  # either copy and paste the URL from a website or download and read in as you would other files. 
df <-
  read_csv("https://open.africa/dataset/36993281-baa5-490d-b6ff-d4750a09d2bb/resource/e7fd9a05-89e2-459a-98ca-e1b40f9a2b1a/download/sub-saharan-africa-ggi-ranks-and-scores.csv",
           skip = 1)  # as the first row is empty (from inspection) skip this

# Read in the iso3 code data
iso <-
  read_csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/refs/heads/master/all/all.csv") |>
  select(1,    # Country name
         3)    # iso3 code
 

```

# Examine the datasets

```{r}

# Check the first and last observations
headTail(df)
headTail(iso)

# Check the structure of the data
glimpse(df)
glimpse(iso)

```

# Special date case

Information contained in the vector `edition` relates to specific years, but these do not contain the necessary info to be converted in a date type vector (e.g., "YYYY-MM-DD"). So, we can leave it as it is. 

# International Organization for Standardization (ISO) labels

You can see that this information is contained in both datasets, but they are labelled differently. The GGGI data labels it `country_iso3`, whereas the iso index labels it `alpha-3`. 

We will need to join these data in order to add the country names to the GGGI data.
This is possible using R's join functions. 


# Join iso information to get country name

```{r}

# Rename alpha-3
iso <-
  iso |>
  rename(country_iso3 = `alpha-3`)

# Join the datasets
df1 <-
  left_join(df, iso, 
            by = "country_iso3") |>
  select(8,1,7,2:6) # re-arrange the vectors to tidy up

# Examine the data
glimpse(df1)

```
We now have names instead of codes only for countries in the dataset.


# Rehape the data

It might actually be useful to split `attribute` into separate ranks and scores.
We can do this using the `pivot_wider` function.

```{r}

# Widen the data by separating attribute
df1 <-
  df1 |>
  pivot_wider(names_from = "attribute",
              values_from = "value")

# Check it worked
head(df1)

```


# Are there any missing values?

Use the `skimr` package function `skim` to generate a report about missing values.

```{r, eval=FALSE}

skim(df1)

```

You can see from the report that there are 8 missing values in the `rank` column. There are no other missin values. 
As we want to plot rank, let's omit these missing values.


# Omit missing values

```{r}

# Use drop_na function to omit rows with missing values
df1 <-
  df1 |>
  drop_na()

```


With the NAs omitted, we can plot the rank data using ggplot2.
Before we do, let's take a look at our vectors again, so we know roughly what our plot hsould look like.

# Data parameters

We want to know, for example, how many observations were recorded per year represented, how many countries, and what the max and min are for rank.
This information will help us identify if something has gone  wrong.


```{r}

# How many years?
df1 |>
  group_by(edition) |>
  count()

```

We can see that the years 2006 to 2024 are included. The number of observations per year (i.e., numbers of countries data) varies too. 

```{r}

# How many countries?
df1 |>
  select(name) |>
  count()

```

The length of the data matches the number of countries (N = 536). It is always worth checking this even if you can infer these details from the data dimensions. 

```{r}

# Min and max for rank
df1 |>
  summarise(max_rank = max(rank),
            min_rank = min(rank))

```

The Max is 151 and the min is 4.


# Plot the data

We now have a rough idea of what the plot will show. 
We are going to plot the GGI index on the y-axis and year on the x-axis.
If we were to plot a line per country the output would be very messy, with 536 unique lines. 
So, let's sample a selection of countries to see how they fair in terms of their GGI ranking.
Ideally, we want countries with data from every year (2006:2024) in the dataset, which we can find out using some more wrangling and summarising.


# Which countries have data for every year in the dataset?

```{r}

# By counting the number of times a name occurs we count the number of years as these are uniqe per row
df1 |>
  count(name) |>
  filter(n ==18) # country will need 18 data points to have a full set (i.e., from 2006-2024)

```

So, from this list let's pick 5: "Botswana", "Chad", "Kenya", "Nigeria" and "Uganda"

```{r}

# Pluck a sample of countries
df2 <-
  df1 |>
  rename(year = edition) |> # change this to year for clarity
  filter(name %in% c("Botswana", 
                     "Chad", 
                     "Kenya", 
                     "Nigeria", 
                     "Uganda"))

# Check the countries included in the data
unique(df2$name)

```

## Plot the data using a line plot

```{r}

# Create a vector for years included in the data
years_incl <-
  seq(2006, 2024, by = 1) # a sequence start with 2006 to 2024, in increments of 1

# Plot the data as a line plot with points
ggplot(df2,
       aes(x = year,
           y = rank,
           fill = name)) +
  geom_line() +
  geom_point(size = 4, 
             shape = 21, 
             alpha = 0.6) +
  labs(x = "\nYear",
       y = "Global Gender Gap Ranking\n",
       fill = "Country") +
  scale_x_continuous(breaks = years_incl) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))

```


# Summary and conclusions

- Today I've shown you how to read data from the web and to tidy and visualise it in R.
- It is important to note that I selected this data because it was non-Western, and the source provided some metadata for me to make sense of it - it's really hard to work with data that you know nothing about!
- As you have learned, programming in R can be challenging and requires you to think carefully about the structure of data.
- However, the pay-off is amazing and worth it in the long run. 
