---
title: "C91AR | Advanced Statistics using R"
subtitle: "Tidying realistic data"
author: "Dr Peter McKenna"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format: 
  revealjs:
    menu: 
      side: right
    smaller: true
    scrollable: true
    slide-number: true
    logo: sossdoc-logo.png
    css: logo.css
    theme: default
editor: 
  markdown: 
    wrap: 72
bibliography: ref.bib
---

```{r include=F, warning=FALSE}

# presentation setup
 knitr::opts_chunk$set(echo = T, 
                       eval = T,
                       results = 'hide',
                       message = FALSE, 
                       warning = FALSE, 
                       tidy.opts = list(width.cutoff = 60),
                       tidy = TRUE)

# default for instructor is 
  # echo = T
  # eval = T
  # results = 'hide'

# you only want to see the results for certain chunks (e.g., plots)

# load packages
library(tidyverse)
library(snakecase)

```


```{r include=F}


#| label: data setup
# load data
df <-
  read_csv("data_raw/robot-expression-data_raw.csv",
           col_types = "ffff?ffffftf")

names(df) <- 
  to_snake_case(names(df)) # lower_case with underlined separators 

# Rename vectors using rename
df <-
  df |>
  rename(perceived_robot_gender = alyx_is, # the new labels are intended to be more descriptive
         robot_expression = expression,
         ppt_acc = resp_acc)

# sort out date col
df <-
  df |> # this is the pipe operator; it means, "then do this"
  mutate(exp_date = dmy(exp_date))

# examine the data
names(df) # notice how the cols are formatted now
 
```


## What we covered in the last session {.smaller}

::: columns
::: {.column width="50%"}
-   Setting up an RProject
-   Creating sub-directories
-   Saving data to a sub-directory
-   Creating an object with that data
:::

::: {.column width="50%"}
-   Setting column data types with the `col_types` argument
-   Generating an overall summary of the data with `summary`
-   Tidy vector labels with `snakecase`
-   Rename vector labels with `rename`
-   Dealing with date vectors using `lubridate`
:::
:::

## Realistic data in research

-   A core strength to using a programming language for data processing
    and analysis is that you reduce the risk of human error
    significantly.
-   This is especially important when confronted with realistic datasets
    that may be untidy for historical or procedural reasons (e.g., a
    database that is maintained by different people who used their
    own unique way of storing and labelling things).
-   An option would be to simply open the data and update the values as
    an when required.
-   However, if the issue is prevalent across many datasets then an
    automated solution is more efficient.

## Today's demonstration

-   We are going to build on what we achieved last session
-   We will be learning a few more tidying operations
-   Then well look at ways to explore and summarise our data

## To get started

-   Open the script you have been working on
-   We are going to continue to work with the
    "robot-expression-data_raw.csv" file
-   To get us back to where we were, run the script in its entirety
-   **Remember** you you only ever need to install packages once (unless
    you are using a different machine) so be sure to comment
    installation commands out

```{r}

# install.packages('tidyverse')

```

## Taking a look at the data

- The `head` function gives you the first 6 rows

```{r results='markup'}

# data summary - head
head(df)

```

---

```{r results='markup'}

# data summary - head specific
  # for specific columns, e.g., cols 1, 5, and 7
head(df[, c(1,5,7)])

```

---

- The `tail` function gives you the last 6 rows

```{r results='markup'}

# data summary - tail
tail(df)


```

---

- The `summary` function summarises continuous and categorical vectors 
- NOTE: `summary` will not count character vectors, that's why we specified that our categorical vectors were factor type (e.g., "f") at `read_csv`

```{r results='markup'}

# data summary - overall
df |>
  summary()

```

---

- You can use the `$` symbol to specify individual cols

```{r results='markup'}

# data summary - sex
df |>
  select(sex) |>
  summary()

```


## Recode variable levels 

- You may have noticed from the last session that there there are some erroneous values in the sex vector.

```{r results = 'markup'}

# Erroneous entries

unique(df$sex) # we can use the `unique` function to check all of the unique values in a vector

# Alternatively you can use tidyverse's `distinct`
# df |>
#   select(sex) |>
#   distinct()

```
-   As you can see, something has gone wrong in the labelling of sex:
    there should only be two values, "Female" and "Male"


## Using recode

-   We can use the `recode` function in combination with `mutate` to
    rename the erroneously named values

```{r}

# label: recode
df <-
  df |>
  mutate(sex = recode(sex,              # mutate to change
                      "male" = "Male",  # supply erroneous value and new value
                      "ffemale" = "Female",
                      "femalew" = "Female"))

# Check the values of a vector using unique
unique(df$sex) # again, we use the $ symbol to isolate a specific vector


```


## Data object considerations

- **NOTE**: If you have assigned transformations to an object with the same original name (e.g., `df` <- `df`) that object is henceforth transformed
- That is why your operations must flow logically from top to bottom, otherwise future operations on the data will not work. 
- The alternative strategy is to give each new object a different name, like `df1`. If you pursue this method, it is worth adding a comment next to each new object specifying how it is different to the previous. 

## Save your tidy data

- Once you are satisfied with the amendments made to your data you can save it as a new tidy `.csv` file
- Again, use comments and bookmarks to signpost this part of your script

```{r eval = FALSE}

# save your new data object into `data_tidy`
write_csv(df, 
          "data_out/robot-expression-data_tidy.csv")

```

## Week 3 Lecture

- Today's lesson is going to include some:
  + recap
  + troubleshooting
  + group activity

# Some important considerations and reflections moving forward

:::columns
:::{.column width=50%}
- **Always** open the `.RProj` file to get the session started, as this will load us onto the correct directory where our data is located.
- When using `<-` and `|>` to modify data objects, either
  + comment out the `<-` line first to check the code works
  + or add `print()` to the end of your pipe chain, then add `<-` once satisfied
:::

:::{.column width=50%}
- We will are working on R scripts - these are the files with the `.R` file type. 
- Make sure you have commented and bookmarked important sections of the script.  
- If you want to work on a data file, you must create an object using `read_csv` and specify the related file directory.
- `|>` used to be represented by `%>%`, but both still work.
:::
:::

# Class activity

- Now that I've guided you through a set of tidying operations on the data, I'd like for you to have a go yourself.
- Start a new R script called `data_tidying.R`
- I'd like for you read in the tidy data file from our `data_tidy` folder and call it `df` to get started.

#  Task

:::columns
:::{.column width=70%}
- The `robot_expression` column labels are meaningless, and need to be recoded.
- Using what we have discussed this far can you recode the values according to the following criteria:
  + **"1" =  "Head up, jaw drop"** 
  + **"2" = "Upper lid raiser, jaw drop"**
  + **"3"=  "Chin raise, head down"**
  + **"4" =  "Eyes closed, head down"**
- Importantly, I'd like you to create a new vector called `rob_head_exp` in the process of doing so
  + **HINT:** you specify the name of the new col at the beginning of the `mutate` argument
:::
  
:::{.column width=30%}  
- I'd also like for you to `select` the following columns
  + id
  + sex
  + eng
  + rob_head_exp
  + ppt_acc
- Write over the tidy file we create last week
:::
:::

# Situations with many input errors

- The above solution is fine when there are only 3 errors in the input values, but what if there are hundreds or thousands?
- You can use the verbs `case_when` and `str_detect` function to identify string patterns (e.g., the starting letter of a string) and replace values based on the detection criteria.

----

```{r eval=FALSE}

# Using case_when and str_detect to modify rows by string patterns
df |>
  mutate(sex = 
           dplyr::case_when(str_detect(sex, 
                                       "^[Mm]") ~ "Male",
                                str_detect(sex, 
                                           "^[Ff]") ~ "Female",
                                TRUE ~ sex))

# Note: ChatGPT is particularly useful for regex patterns

```

- This might look like a foreign language right now, so let me parse it out for you

# Parsing out the use of `case_when` and `str_detect`

::: columns
::: {.column width="50%"}
-   `case_when` is like `recode`, only it also requires a conditional
    statement (or an additional condition) in its application.
-   In our case, we want R to look for values that start and end with
    upper and lower case "m" and "f" to to give consistency to the
    "Male" and "Female" labelling in our `sex` vector.
:::

::: {.column width="50%"}
-   `str_detect` is a function from the **stringr** package that allows
    you to perform operations on strings in a vector.
-   In this case, we have asked `str_detect` to look for cell values
    that start with an upper or lower case "m" and "f" in the sex
    vector.
-   We then supply the alternate value after the `~` in `case_when`
-   We finish the argument chain by stating that the rest of the values,
    if not matched, come from the sex vector.
:::
:::

---

# Making use of our new tidy object

- Now that we are happy with our tidied dataset we can start a new script focussed on summarising and plotting the data.
- Before you do though, have a look through your current script, run from start to end to make sure it works.
- Make sure the code is tidy and that you have commented efficiently.
- Once you are happy, start a new script called `data_wrangle`

# Tidyverse verbs

- `mutate()` adds new variables that are functions of existing variables
- `select()` picks variables based on their names.
- `filter()` picks cases based on their values.
- `summarise()` reduces multiple values down to a single summary.
- `arrange()` changes the ordering of the rows.
- [More information about Tidyverse verbs here](https://dplyr.tidyverse.org/)

# Reading in data for wrangling

- You've probably figured out what we are going to do next
  + read in the tidy dataset into our new script
- This is my preferred workflow in R; to separate *tidying* from *wrangling*
- Note, when I say *wrangling* here, I'm referring to all the operations I might perform on tidy data, e.g., summarising, plotting, analysing

```{r eval = F}

# Wrangling data

# load packages
library(tidyverse) # load in our toolkit

# read in data
df <-
  read_csv("data_out/robot-expression-data_tidy.csv") # don't forget the `.csv` bit!


```

# Filtering data

- Another important operation you will want to do is to filter data depending on a set criteria
- This is easily achieved using the `filter` function
- With filter you specify the vector and value of interest. These must follow the same convention as the data (e.g., labels must match)
- Let's say we only want to look at males whose first language was English

```{r}

df |>
  filter(sex == "Male", # this only works because we have a vector called `sex` which has values labelled `Male`
         eng == "Yes") |>
  print()

```


# `filter` activity

- Can you filter an object based on this criteria
  + **Females**
  + **Asian ethnicity**
  + **Never interacted with a robot before**
- **HINT:** You can examine the names of the levels of a vector using the `unique` function  

## Using `select` to subset data

-   the `select` verb allows you to both select and rename columns (if
    you wish to)
-   You can refer to columns by their name (e.g., "id") or their
    position on the dataframe (e.g., "1")
-   Let's subset the data by native language and response accuracy

```{r}

# label: select

# Subset by native language and response accuracy 
df |>
  select(eng,
         ppt_acc) |>
  print()

```

---

- Sub-setting and renaming can be done as follows

```{r}

# label: select and rename

# Subset by native language and response accuracy 
df |>
  select(native_english = eng,
         participant_accuracy = ppt_acc) |>
  print()

```

# A glimpse into summary chains

- Summarise count of response accuracy by robot expression.

```{r}

df |>
  group_by(robot_expression,
           ppt_acc) |>
  count()

```

# C91AR | Tidying realistic data | Lab

## Recap on what we've covered so far

::: columns

:::{.column width="50%"}
- setting up a project
- saving data in our directory
- reading in data and specify `col_types`
- standardising vector level formatting with `snakecase`
- changing vector labels with `rename`

:::

:::{.column width="50%"}
- reformatting date vectors with `lubridate`
- explore data 
- recode levels of a vector
- filtering data
- sub-setting by specific vectors
- saving data into a new .csv after tidying it up

:::
:::

# Before we move onto the next thing today is there anything you would like me to revisit?

## Week 3 lab objectives

- Apply what you've learned to a new dataset
  + read in the data
  + specify column types
  + tidy vector names
  + rename where appropriate
  + recode levels of a vector
  + filter data
  + select specific columns
  + saving data into a new .csv after tidying it up




