---
title: "My first Quarto doc"
author: "Peter McKenna"
format: html
editor: visual
---

# Learning about Correlation through Data Simulation

```{r warning=FALSE}

# Load packages
library(tidyverse)
library(corrr)
library(psych)
library(GGally)

# Read in the data
handw <-
  read_csv("data_raw/heights_and_weights.csv")
```

## Examine the data

```{r}

# Examine the data
glimpse(handw)
head(handw)
tail(handw)
```

The heights and weights dataset contains 475 observations and 2 vectors: `height_in` and `weight_lbs`.

## Descriptive Statistics

```{r}

# Gather means and SDs of handw
describe(handw)

```

The mean of height in inches was 62.91 (SD=12.12). The mean for weight in pounds was 132.28 (SD=54.19). Measures of both skew (-1.78) and kurtosis (2.89) indicate that height in inches was not normally distributed.

## Plot the distribution of height in inches

```{r}

# Generate a histrogram of height in inches
ggplot(data = handw,
       mapping = aes(x = height_in)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Histrogram of the distribution of height in inches",
       x = "\nHeight in inches",
       y = "Frequency\n")

```

## Copying and pasting code

```{r}
tibble(
  height_cm = rnorm(30, mean = 179, sd = 13),   # generate data on a normal distribution
  weight_kg = rnorm(30, mean = 78, sd = 5)) %>%
  mutate_if(is.numeric, round, digits = 0) %>%  # remove decimals for simplicity
  mutate(id = row_number())                     # add an id column based on the row number  
```
