---
title: "Computer Lab 8: Regression modelling"
subtitle: "C91AR: Advanced Statistics using R"
author: "Dr Peter E McKenna"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
bibliography: ref.bib
output: 
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    includes:
      in_header: preamble.tex
    keep_tex: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

# markdown formatting
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

```

# Setup code

```{r}

# change output format
options(scipen = 999)

# set the seed
set.seed(453)

# load packages
pacman::p_load(corrr,
               tidyverse,
               psych,
               tidyplots)

```

# Outline for today's session

Using the slides from Monday..

- develop a regression model for for either the 
  - "fathers_son_height.csv"
  - "mother_daughter_height.csv"
- Use the dataset **that you did NOT simulate last Thursday**
- Use mathematics notation in your report


# Futher instructions

  - see [my Markdown doc for Monday's lecture on Github](https://github.com/petermckenna/MSc_C91AR/blob/main/L7_Correlation_and_regression_pdf.rmd) to familiarise yourself with the notation format using $ wrappers. These are to be written in the regular text sections of your Markdown document. 
- Check your model against the results of running the model using `lm(DV ~ IV, data = data)`   
- Once you have created your model, predict a new value for
  - a parent's height based on their child's height
  - a child's height based on their parent's height 
- Visualise the data with a regression line to check that your predicted values fall within the model's parameters.
  
# As always

- Annotate and comment your report
- Ask for help if you need it
- Take. Your. Time. 

  
  
  

